Microxanox - an R package for simulating an $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states.

### 2022-04-26 inst/manuscript/manuscript_microxanox.Rmd
Update manuscript_microxanox.Rmd

```{r setup, include = FALSE}
library(plantuml)
```

@@

@@

# Motivation and significance
Many ecosystems are exposed to gradual changes of environmental variables, to which the responses are not always as gradual as the change of the environmental variable. One example where the gradual change of a single environmental causing an abrupt change of the system is the switch from an aerobic to anaerobic system. This system hass been investigated by @Bush2017. We wanted to take this investigation one step further, and look at the role of an increased biodiversity plays in these dynamics [@REF_NEEDED]. For this purpose, we developed this package.

The microxanox package is a package for simulating a three functional group system (*CB*: cyanobacteria, *PB*: phototrophic sulfur bacteria, and *SB*: sulfate-reducing bacteria) with four chemical substrates (*P*: phosphorus, *O*: oxygen, *SR*: reduced sulfur, *SO*: oxidized sulfur). It includes feddback between biogeochemical processes and is based on [@Bush2017] (See [@Bush2017] for a detailed discussion of the model).

The aims of the `microxanox` package are twofold. Firstly, the package aims at reproducing the results shown by [@Bush2017], which is acomplished in the [vignette Partial reproduction of Bush et al](LINK NEEDED). Secondly, to take these results one step further, it includes new functionality to adress our research question as presented in [@REF_NEEDED].

@@

In addition to the model itself, the package includes some functions to analyse the results as well as visualise the results to provide a starting point for customised visualisations based on own requirements.

# Software description
{>>Describe the software in as much as is necessary to establish a vocabulary needed to explain its impact.<<}

```{plantuml runsim_example, eval = TRUE, plantuml.format = "auto", plantuml.preview = TRUE}
@startuml
partition "Run Single Simulation" {
   start
  :parameter <- new_runsim_parameter();
  split
    #palegreen:saveRDS(FILE, parameter);
    detach
  split again
    #pink:result <- run_simulation(parameter);
  end split
  split
    #palegreen:saveRDS(FILE, results);
    detach
  split again
    :Change some parameter;
    #pink:result2 <- run_simulation(results);
  end split
  split
    #palegreen:saveRDS(FILE, results2);
    detach
  split again
    end
  end split
}
@enduml
```

The package is not intended to provide a modelling framework which can be adjusted easily to all needs, but primarily a tool to implement the model used by [@Bush2017] and to extend it to our needs [@REF_NEEDED]. Consequently, any more substantial changes and adaptaitions, are likely to need a change in the source code.

Nevertheless, the model is structured in a way which builds on a modular structuere, so that e.g. the event definition can easily be changed. or other aspects can be adjusted. All values in the parameter object can be changed as needed and the general structure of the code should make it not to difficult to adapt the model to other similar systems.

## Software Architecture
{>>Give a short overview of the overall software architecture; provide a pic- torial component overview or similar (if possible). If necessary provide im- plementation details.<<}

The framework used when writing this package aims at reproducibility of the results. It builds on the folowing main considerations:

@@

A typical simulation would look as followed:

```{r, eval = FALSE}
## Create the parameter
parameter <- new_runsim_parameter()
# manually setting certain parameter

## Run the simulation and save the result
result <- run_simulation(parameter)
saveRDS(result, "sim1.rds")

## Do other stuff, e.g. plotting

## Load results, change some parameter, and rerun the simulation and save the result
parameter <- loadRDS("sim1.rds")
# change some parameter
result <- run_simulation(parameter)
saveRDS(result, "sim2,rds")
```

## Software Functionalities
{>>Present the major functionalities of the software.<<}

We will now discuss the general structure and functionality of the package without going into to much detail. A more detailed discussion can be found in the [User Guide](LINK_NEEDED).

@@

# Illustrative Examples
{>>Present the major functionalities of the software.<<}

### Running one simulation

The individual simulation (`run_simulation()` function) is the working horse in this package. In this function, the ODEs are solved. The function needs only one argument - an object as created by the function `new_runsim_parameter()`. One parameter of this object is the `strain_parameter` which can be created by the function `new_strain_parameter()`. For a detailed description of the parameter and how they are created please see the User Guide and which acomnpanies the package or is available at [User Guide](@LINK_NEEDED)

@@

```{r plot-dynamics, echo = FALSE,  include=TRUE, fig.cap='Plot of results of a simulation run using the function $plot_dynamics()$. Details can be found in the "User Guide" section "Three strains per functional group".', fig.align="center", out.width = 500 }
knitr::include_graphics("figures/ug_three_strains_dynamics.png")
```

### Finding a Steady State of the model

@@

#### Replication Method
The replication method is implementad in the function `run_replication_ssfind()` which takes a parameter object as returned by the function `new_replication_ssfind_parameter()` and the number of cores for multithreading the simulation.

#### Temporal Method
The temporal method involves two simulations for a particular system configuration (parameter set). In one simulation the oxygen diffusivity is *increased* in a step-wise fashion. In the other it is *decreased* in a step-wise fashion. That is, oxygen diffusivity is held at a constant level for long enough for steady state to be reach, that state is recorded, and then a slightly higher (or lower) oxygen diffusivity value is set. Hence, at that time point, the system is effectively started with initial conditions that are the state of the system in the previous time step.

@@

### Extracting Stability Measures
From the raw results returned by these `run_...()` functions, the stability measures can be extracted by using the function `get_stability_measures()`. These measures include non-linearity and hysteresis measures, of the response of the simulated system to environmental change.

# Impact
{>>This is the main section of the article and the reviewers weight the description here appropriately
Indicate in what way new research questions can be pursued as a result of the software (if any).
Indicate in what way, and to what extent, the pursuit of existing research questions is improved (if so).
Indicate in what way the software has changed the daily practice of its users (if so).
Indicate how widespread the use of the software is within and outside the intended user group.
Indicate in what way the software is used in commercial settings and/or how it led to the creation of spin-off companies (if so).<<}

@@

```{r uc1_stable_state, echo = FALSE,  include=TRUE, fig.cap='Plot of the stable states of the simulation runs under different oxygen diffusivity. The top graph are the Organisms (ech initially with three strains) while the lower graph is the substrate availability under the same oxygen diffusivities. Details can be found in the "User Guide" section "Three strains per functional group".', fig.align="center", out.width = 500 }
knitr::include_graphics("figures/ug_three_strains_stable_state.png")
```

@@

```{r user_guide_hysteresis, echo = FALSE,  include=TRUE, fig.cap='Hysteresis of all assessed combinations of variability.', fig.align="center", out.width = 500 }
knitr::include_graphics("figures/user_guide_hysteresis.png")
```

>>>>

```{r setup, include = FALSE}
```

@@

# TODO
- **update links**
- **update crossref**

@@

# Motivation and significance
Many ecosystems are exposed to gradual changes of environmental variables, to which the responses are not always as gradual as the change of the environmental variable. One example where the gradual change of a single environmental causing an abrupt change of the system is the switch from an aerobic to anaerobic system. This system has been investigated by @Bush2017. We wanted to take this investigation one step further, and look at the role an increased biodiversity plays in these dynamics [@REF_NEEDED]. For this purpose, we developed this package.

The microxanox package is a package for simulating a three functional group system (*CB*: cyanobacteria, *PB*: phototrophic sulfur bacteria, and *SB*: sulfate-reducing bacteria) with four chemical substrates (*P*: phosphorus, *O*: oxygen, *SR*: reduced sulfur, *SO*: oxidized sulfur). It includes feedback between biogeochemical processes and is based on [@Bush2017] (See [@Bush2017] for a detailed discussion of the model).

The aims of the `microxanox` package are twofold. Firstly, the package aims at reproducing the results shown by [@Bush2017], which is accomplished in the [vignette Partial reproduction of Bush et al](LINK NEEDED). Secondly, to take these results one step further, it includes new functionality to address our research question as presented in [@REF_NEEDED].

@@

In addition to the model itself, the package includes some functions to analyse the results as well as visualize the results to provide a starting point for customized visualizations based on own requirements.

# Software description
> {>>Describe the software in as much as is necessary to establish a vocabulary needed to explain its impact.<<}

The package is not intended to provide a modelling framework which can be adjusted easily to all needs, but primarily a tool to implement the model used by [@Bush2017] and to extend it to our needs [@REF_NEEDED]. Consequently, any more substantial changes and adaptations, are likely to need a change in the source code.

Nevertheless, the model is structured in a way which builds on a modular structure, so that e.g. the event definition can easily be changed, or other aspects can be adjusted. All values in the parameter object can be changed as needed and the general structure of the code should make it not to difficult to adapt the model to other similar systems.

## Software Architecture
> {>>Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide im- plementation details.<<}

The framework used when writing this package aims at reproducibility of the results. It builds on the following main considerations:

@@

## Software Functionalities
> {>>Present the major functionalities of the software.<<}

We will discuss the general structure and functionality of the package without going into to much detail. A more detailed discussion can be found in the [User Guide](LINK_NEEDED).

@@

# Illustrative Examples
> {>>Present the major functionalities of the software.<<}

## Running one simulation

A typical simulation would look as shown in \@ref(fig:runsim_example).

```{R runsim_example, eval = TRUE, echo = FALSE, fig.cap='Typical flow of a simulation. Dark Grey boxes: commands necessary for simulation; Light Grey:Saving of parameter and results; Lightest Grey: Different non specified commands.', fig.align="center", out.width = 350 }
fig <- "./figures/simflow.pdf"

if (!file.exists(fig)){
  library(plantuml)
  puml <- '
    @startuml
    partition "Run Single Simulation" {
      start
      #DarkGrey:\"\"library(microxanox)\"\"\n\"\"parameter <- new_runsim_parameter(...)\"\";
      :Change some parameter
      in the \"\"parameter\"\" object;      
      split
        #LightGrey:\"\"saveRDS(FILE, parameter)\"\";
        detach
      split again
        #DarkGrey:\"\"result <- run_simulation(parameter)\"\";
      end split
      split
        #LightGrey:\"\"saveRDS(FILE, results)\"\";
        detach
      split again
        #DarkGrey:\"\"plot_dynamics(results)\"\";
        :Change some parameter
        in the \"\"result\"\" object;
        #DarkGrey:\"\"result2 <- run_simulation(results)\"\";
      end split
      split
        #LightGrey:\"\"saveRDS(FILE, results2)\"\";
        detach
      split again
        #DarkGrey:\"\"plot_dynamics(results)\"\";
        end
      end split
    }
    @enduml
  '

  plantuml::get_graph(puml, fig)
}

knitr::include_graphics(fig)
```

The individual simulation (`run_simulation()` function) is the working horse in this package. In this function, the ODEs are solved. The function needs only one argument - an object as created by the function `new_runsim_parameter()`. One parameter of this object is the `strain_parameter` which can be created by the function `new_strain_parameter()`. For a detailed description of the parameter and how they are created please see the User Guide which acomnpanies the package or is available at [User Guide](@LINK_NEEDED)

@@

```{r plot-dynamics, echo = FALSE,  include=TRUE, fig.cap='Plot of results of a simulation run using the function $plot_dynamics()$. Details can be found in the "User Guide" section "Three strains per functional group".', fig.align="center", out.width = 350 }
knitr::include_graphics("figures/ug_three_strains_dynamics.png")
```

## Finding a Steady State of the model

@@

### Replication Method
The replication method is implementad in the function `run_replication_ssfind()` which takes a parameter object as returned by the function `new_replication_ssfind_parameter()` and the number of cores for multithreading the simulation.

### Temporal Method
The temporal method involves two simulations for a particular system configuration (parameter set). In one simulation the oxygen diffusivity is *increased* in a step-wise fashion. In the other it is *decreased* in a step-wise fashion. That is, oxygen diffusivity is held at a constant level for long enough for steady state to be reach, that state is recorded, and then a slightly higher (or lower) oxygen diffusivity value is set. Hence, at that time point, the system is effectively started with initial conditions that are the state of the system in the previous time step.

@@

## Extracting Stability Measures
From the raw results returned by these `run_...()` functions, the stability measures can be extracted by using the function `get_stability_measures()`. These measures include non-linearity and hysteresis measures, of the response of the simulated system to environmental change.

# Impact
> {>>This is the main section of the article and the reviewers weight the description here appropriately
Indicate in what way new research questions can be pursued as a result of the software (if any).
Indicate in what way, and to what extent, the pursuit of existing research questions is improved (if so).
Indicate in what way the software has changed the daily practice of its users (if so).
Indicate how widespread the use of the software is within and outside the intended user group.
Indicate in what way the software is used in commercial settings and/or how it led to the creation of spin-off companies (if so).<<}

@@

```{r uc1_stable_state, echo = FALSE,  include=TRUE, fig.cap='Plot of the stable states of the simulation runs under different oxygen diffusivity. The top graph are the Organisms (ech initially with three strains) while the lower graph is the substrate availability under the same oxygen diffusivities. Details can be found in the "User Guide" section "Three strains per functional group".', fig.align="center", out.width = 350 }
knitr::include_graphics("figures/ug_three_strains_stable_state.png")
```

@@

```{r user_guide_hysteresis, echo = FALSE,  include=TRUE, fig.cap='Hysteresis of all assessed combinations of variability.', fig.align="center", out.width = 350 }
knitr::include_graphics("figures/user_guide_hysteresis.png")
```
